---
title: "Univariate Time Series Analysis: Assignment 1"
subtitle: "HS 2026, University of Basel"
author:
  - "First Name Last Name (team member 1)"
  - "First Name Last Name (team member 2)"
  - "First Name Last Name (team member 3)"
date: today
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
template for your submission. replace the names above (up to three team members),
fill in the code chunks and write your answers in full sentences below each question.
this file must render without errors when it sits in the same folder as
StarbucksSales.csv. load every package you use in the first chunk.
-->

```{r}
# packages


# data (relative path, file next to this .qmd)
starbucks <- read.csv("StarbucksSales.csv", sep = ";")
```

# Problem 1 (preprocessing and data analysis)

> we analyse the quarterly sales of Starbucks Corporation (in million USD) from January 1995 to December 2018. the data `StarbucksSales.csv` are on OLAT; the file runs to December 2024, use only the observations up to December 2018. provide your R code, present your results clearly and justify your answers where required.

## (a)

> load the data set and plot the time series from January 1995 to December 2018. make sure the plot is clear. make three observations from the graph.

```{r}

```

**Observations:**

1. ...
2. ...
3. ...

## (b)

> find the two highest sales values and give the corresponding year and quarter.

```{r}

```

**Answer:** ...

## (c)

> consider the regression model with a linear trend
> $$
> \mathit{Sales}_t = \beta_0 + \beta_1 \cdot t + \epsilon_t.
> $$
> estimate the model by ordinary least squares, present the results, and add the estimated trend to the graph from (a).

```{r}

```

**Answer:** ...

## (d)

> use `predict()` to compute the predicted values for the four quarters of 2019.

```{r}

```

**Answer:** ...

## (e)

> we now apply the classical decomposition of a time series with linear filters. all steps must be computed manually, i.e. R functions such as `decompose()` are not allowed. first, estimate the trend $T_t$ with the one-sided simple moving average over the past four quarters, $T_t := \frac{1}{4} \sum_{j=0}^{3} \mathit{Sales}_{t-j}$, and plot $T_t$.

```{r}

```

**Answer:** ...

## (f)

> building on (e), estimate the seasonal component $S_t$ and the remainder $E_t$ and plot them in two separate graphs. what do you observe in these two plots?

```{r}

```

**Observations:** ...
