---
title: "Exercise: Market price with heterogeneous beliefs and risk aversion"
subtitle: "Predictions: Theory and Practice"
author: "Patrick Schmidt"
institute: "University of Heidelberg"
params:
  show_solutions: true
format:
  html:
    toc: true
    toc-depth: 3
    self-contained: true
    theme: cosmo
  pdf:
    documentclass: article
    geometry: margin=2.5cm
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
theme_set(theme_bw() + theme(text = element_text(size = 14)))
```

## Setting

An asset trades today at price $S$ and pays a random amount $Y$.

There are **two traders**, $A$ and $B$, each choosing a position $q_i \in \mathbb{R}$:

- $q_i > 0$: trader $i$ buys asset (pays $S$ today, receives $Y$ tomorrow)
- $q_i < 0$: trader $i$ sells asset

Each trader has **mean-variance preferences** over the position payoff $q_i(Y - S)$:

$$
U_i(q_i) \;=\; q_i \,(\mu_i - S) \;-\; \frac{\gamma_i}{2}\, q_i^{\,2}\, \sigma^2,
$$

where

- $\mu_i = E_i[Y]$ is trader $i$'s **mean belief** about $Y$,
- $\sigma^2$ is the (common) belief about the variance of $Y$,
- $\gamma_i > 0$ is trader $i$'s **risk aversion**.

$A$ expects higher returns: $\mu_A > \mu_B$.

We will look for the market price $S$ in which $A$ wants to buy and $B$ wants to sell the same amount $q_i$, i.e. the market clearing price.

You can assume that $S$ lies between $\mu_B$ and $\mu_A$.


### (a) Supply by the pessimist

For a given price $S > \mu_B$, derive trader $B$'s optimal position $q_B^*(S)$ and the resulting **supply** $q_B^*(S)$ to the market.

How does the supply curve depend on $S$, $\gamma_B$, $\sigma^2$, and $\mu_B$?

::: {.content-visible when-meta="params.show_solutions"}
#### Solution

The first-order condition for $B$ is
$$\frac{\partial U_B}{\partial q_B} \;=\; (\mu_B - S) - \gamma_B\, q_B\, \sigma^2 \;=\; 0
\;\Longrightarrow\; q_B^*(S) \;=\; \frac{\mu_B - S}{\gamma_B\, \sigma^2}.$$

For $S > \mu_B$ we have $q_B^* < 0$: trader $B$ goes short. The **supply** to the market is
$$\boxed{\;\text{supply}(S) \;=\; -q_B^*(S) \;=\; \frac{S - \mu_B}{\gamma_B\, \sigma^2}\;}$$

Properties:

- **decreasing** in risk aversion $\gamma_B$: more risk-averse pessimist supplies less
- **decreasing** in variance $\sigma^2$: more uncertainty → smaller short position
- **decreasing** in $\mu_B$: a more pessimistic trader is selling more at any given $S$
:::


### (b) Demand by the optimist

For a given price $S < \mu_A$, derive trader $A$'s optimal position $q_A^*(S)$, which gives the **demand** in the market.

How does the demand curve depend on $S$, $\gamma_A$, $\sigma^2$, and $\mu_A$?

::: {.content-visible when-meta="params.show_solutions"}
#### Solution

By symmetry with (a), the FOC for $A$ gives
$$\boxed{\;\text{demand}(S) \;=\; q_A^*(S) \;=\; \frac{\mu_A - S}{\gamma_A\, \sigma^2}\;}$$

For $S < \mu_A$ this is positive: trader $A$ buys.

Properties:

- linear and **decreasing** in $S$
- **decreasing** in risk aversion $\gamma_A$: more risk-averse optimist buys less
- **decreasing** in variance $\sigma^2$
- **increasing** in $\mu_A$: the more optimistic the trader, the more aggressively she buys
:::


### (c) Equilibrium price

Solve for the equilibrium price $S^*$ where demand equals supply.

::: {.content-visible when-meta="params.show_solutions"}
#### Solution

Set demand equal to supply:
$$\frac{\mu_A - S^*}{\gamma_A\, \sigma^2} \;=\; \frac{S^* - \mu_B}{\gamma_B\, \sigma^2}.$$

The $\sigma^2$ cancels. Cross-multiplying:
$$\gamma_B\,(\mu_A - S^*) \;=\; \gamma_A\,(S^* - \mu_B)$$
$$\gamma_B\, \mu_A + \gamma_A\, \mu_B \;=\; (\gamma_A + \gamma_B)\, S^*$$

$$\boxed{\; S^* \;=\; \frac{\gamma_B\, \mu_A + \gamma_A\, \mu_B}{\gamma_A + \gamma_B} \;}$$

- $S^*$ is a **weighted average** of the two beliefs, where each trader's weight is inversely proportional to their own risk aversion:
$$S^* \;=\; w_A\, \mu_A + w_B\, \mu_B, \qquad w_i \;=\; \frac{1/\gamma_i}{1/\gamma_A + 1/\gamma_B}.$$

- The **less risk-averse** trader has **more influence** on the price.

**Special cases.**

- Equal risk aversion $\gamma_A = \gamma_B$: $\quad S^* = \tfrac{1}{2}(\mu_A + \mu_B)$ 

  — simple average of beliefs.
  
- Trader $A$ becomes risk-neutral ($\gamma_A \to 0$): $\quad S^* \to \mu_A$ 

  — the aggressive trader's belief dominates.
  
- Same belief $\mu_A = \mu_B$: $\quad S^* = \mu_A = \mu_B$ 

  — no disagreement, no trade.

Note that $\sigma^2$ does **not** appear in $S^*$: 

- variance scales demand and supply by the same factor, so it cancels in equilibrium. 

- It does affect the **traded quantity** (less variance → larger $q_A^* = -q_B^*$).

:::


### (d) Numerical example

Take $\mu_A = 110$, $\mu_B = 90$, $\gamma_A = 1$, $\gamma_B = 2$, $\sigma^2 = 100$.

- Compute $S^*$ using the formula from (c).
- The plot below shows demand and supply.

```{r demand_supply_plot, echo=FALSE, fig.height=4, fig.width=6, fig.align="center"}
mu_A <- 110; mu_B <- 90
gA   <- 1;   gB   <- 2
sig2 <- 100

S_eq <- (gB * mu_A + gA * mu_B) / (gA + gB)

S_grid <- seq(85, 115, length.out = 200)
dat <- tibble(
  S      = S_grid,
  demand = pmax(0, (mu_A - S) / (gA * sig2)),
  supply = pmax(0, (S - mu_B) / (gB * sig2))
)

ggplot(dat, aes(x = S)) +
  geom_line(aes(y = demand, colour = "demand"), linewidth = 1.1) +
  geom_line(aes(y = supply, colour = "supply"), linewidth = 1.1) +
  geom_vline(xintercept = S_eq, linetype = "dashed") +
  annotate("text", x = S_eq + 0.3, y = 0.18,
           label = paste0("S* = ", round(S_eq, 2)), hjust = 0) +
  scale_colour_manual(values = c(demand = "steelblue", supply = "firebrick")) +
  labs(x = "price S", y = "quantity", colour = NULL) +
  theme(legend.position = "bottom")
```

Check that the formula matches the intersection on the plot.

::: {.content-visible when-meta="params.show_solutions"}
#### Solution

$$S^* \;=\; \frac{\gamma_B\, \mu_A + \gamma_A\, \mu_B}{\gamma_A + \gamma_B} \;=\; \frac{2 \cdot 110 + 1 \cdot 90}{1 + 2} \;=\; \frac{310}{3} \;\approx\; 103.33.$$

The optimist $A$ carries weight $w_A = (1/1)\,/\,(1/1 + 1/2) = 2/3$, the pessimist $B$ carries $1/3$. The price sits closer to $A$'s belief because $A$ is less risk-averse.

Equilibrium traded quantity:
$$q_A^*(S^*) \;=\; \frac{\mu_A - S^*}{\gamma_A\, \sigma^2} \;=\; \frac{110 - 103.33}{1 \cdot 100} \;\approx\; 0.067.$$
$A$ goes long $\approx 0.067$ units; $B$ goes short the same amount.
:::


### (e) Connection to the lecture

In the lecture we wrote that the market price equals the **median** of the beliefs (or a wealth-weighted version of it). Here we obtained a **mean** of the beliefs. Explain in one or two sentences how the two statements are compatible.

::: {.content-visible when-meta="params.show_solutions"}
#### Solution

Both statements say "the equilibrium price aggregates heterogeneous beliefs and sits *between* the optimists and pessimists." Which exact statistic appears depends on the model:

- with **continuous** positions and mean-variance / CARA-normal preferences (this exercise) → **risk-tolerance-weighted mean** of beliefs
- with **bounded** positions (e.g. a prediction-market contract that can be held at most once) and risk-neutral, equally wealthy traders → **median** of beliefs
- with heterogeneous wealth → wealth-weighted version of either

In all variants, traders who are **less risk-averse, wealthier, or face fewer position limits** have **more influence** on the price.
:::


## Aside: mean-variance utility

The objective
$$U_i(q_i) \;=\; q_i(\mu_i - S) \;-\; \tfrac{\gamma_i}{2}\, q_i^{\,2}\, \sigma^2$$
arises from CARA utility $u_i(W) = -\exp(-\gamma_i W)$ over final wealth, plus the assumption that $Y \sim \mathcal{N}(\mu_i, \sigma^2)$ under trader $i$'s belief. Under CARA + normal, expected utility is monotone in $E_i[W] - (\gamma_i / 2)\,\mathrm{Var}_i(W)$, i.e. the mean-variance form above.
