---
title: "Predictions: Theory and Practice"
subtitle: "Part III, Computational Methods"
---
```{r setup, echo=F,message=F}
library(ggplot2)
theme_set(theme_bw()+theme(text = element_text(size=16)))

knitr::opts_chunk$set(eval = TRUE, echo = TRUE,message = FALSE,warning = FALSE)
```


# Introduction

## Software

- in the lecture, I will mostly use R

    - some solutions will also be available in Python
    
    - I use quarto/markdown documents in RStudio, which can execute R and Python examples
    

### List of tools

- R / python are open-source and free programming languages

- you can run both on your laptop (after installing them)

  - **IDE** help in using them (+AI help)
  
  - for R: **RStudio**
  
  - for python: Visual Studio Code, Jupyter, PyCharm

- you can run both only **web-based** (less preparation needed)

  - for R: Kaggle, Posit cloud, Google Colab
  
  - for python: Kaggle, Jupyter Notebook, Google Colab



---

- for Bayesian modeling we will use 

    - Stan (R,python), JAGS(R)
    
    - and pymc (python)

- For Stan, there exists easy to use web-implementation

    - https://stan-playground.flatironinstitute.org/
    

### Programming

- I will try to provide context during the lecture

- some learning by doing will be necessary

- I recommend spending a few hours on an intro/tutorial to R (or python)

- great starting point: LLMs

    - "I want to do XXX in R/python"
    
    - "explain what the following code does"
    
    - "I have the following code and error"

### Agentic programming

- an LLM put inside a software "harness" that lets it act, not just chat

  - "tools" = functions the harness provides for the llm 

  - it can **read** your files, **write** and **run** code

  - you describe a goal in plain language; the agent plans and executes the steps

- the loop: **think, questions, plan, execute, test (and fix)**

- probably how a lot of programming will be done in the future

### CLI-based vs. other agentic tools


- **Web / chat-based** (in the browser, sandboxed)

  - safest and easiest to start

  - examples: ChatGPT, Claude.ai, Gemini, Google AI Studio
  
  - harness lives online
    
    - that is how they search the web, etc.

- **IDE / editor-integrated** (live inside your code editor)

  - common IDEs with addins: Visual Studio Code, RStudio
  
  - specific IDEs: Cursor (free student tier)

  - suggestions and chat next to your code

---

- **Command Line (CLI) / terminal-based agents** (run in a terminal, full access to your machine)

  - operate directly in CLI on your folders, run any command

  - most powerful and flexible

  - examples: Claude Code, Codex CLI, Antigravity CLI, OpenCode (needs API for llm)
  
  - runs in the background for Claude Cowork and Codex

### What is the right workflow for you?

- you: no interest in coding.

  - use web-based interfaces
  
  - understand logic of method and interpretation of output
  
- you: some interest

  - python, cloud: use google colab with integrated Gemini 3 flash
  
  - python, local: Antigravity or Visual Studio Code
  
  - R, local: Rstudio with Antigravity Cli in background or just copy paste 

- you: highly interested

  - many options, local installation still necessary
  
  - combination of IDE and Cli tool
  
  - OpenCode Cli with free tier or Openrouter free models
  
  - some setup costs

### Paid subscriptions?

- not necessary or helpful for this lecture

- not necessary for learning the workflow of agentic coding

- but very helpful for implementing empirical work that would be based on this lecture

- unclear how this develops

  - free tier catching up and being smart enough for student projects?
  
  - paid tier usage forbidden for graded work?
  
  - student access via university (nothing helpful planned at this point)?
  
  


## Overview

- Monte-Carlo simulations

    - generate probabilistic models that can quantify uncertainty
    - input: data generating process
    
- Bayesian modeling

    - generate probabilistic models that can learn from data by Bayesian updating
    - input: likelihood function + data
    - learn MCMC sampling
    - priors in applications

- Machine learning

    - learn methods that scale better than Bayesian learning
    - test, train, validate
    - loss functions
    - deep learning basics


# Model-based predictions with Monte-Carlo simulations

## Introduction

- we want to predict an uncertain outcome in a complex system

- the prediction should be probabilistic

- we consider a situation without data

- Approach: Monte-carlo simulation



### Examples

* Example Questions:
    
    * How much does a stock market investment return after 10 years?
    
    * How often do I win in the Monty-Hall / Goat problem if I switch doors?


### Monte Carlo Simulation: The Core Idea

* **Analogy:** Imagine rolling dice thousands of times to figure out the probability of getting different numbers or combinations. 

    - Monte Carlo does this computationally for complex systems.

* **Core Idea:** Use repeated **random sampling** from input probability distributions to simulate the behavior of a system and generate a distribution of possible outcomes.

* As the **number of simulation repetition increases**, the empirical distribution **converges towards the true distribution**

* Named after the Monte Carlo Casino

## Approach

### How It Works: The Steps

1.  **Model the System:** Define the mathematical relationships or rules that link inputs to outputs.

2.  **Assign Probability Distributions:** For each uncertain input, define its probability distribution 
    
    - based on historical data, expert opinion, or assumptions
    
3.  **Generate Random Samples:** One simulation run (an "iteration" or "trial"):
    
    - draw a random value for *each* uncertain input
    
    - calculate outcome based on model

4.  **Repeat:** Perform step 3 many times (number of iterations / simulation runs)

5.  **Aggregate Results:** Collect all the outcomes from the iterations which are a sample of the underlying distribution

6. **Visualize** or **Summarize** this sample

    - event outcome: compute estimated probability by ratio
    
    - real outcome: histograms for the distribution and average for mean


### Strengths

* **Quantifies Uncertainty:** 

    - provides a full distribution of possible outcomes
    
* **Handles Complexity:** 

  - can model complex systemns
  
* **Intuitive:** 

  - relatively easy to grasp
  
* **Low requirements:** 

  - needs only a generative model / data generating process



### Weaknesses

- cannot include data

- **"Garbage In, Garbage Out":** 

    - results depend on model choice and assigned probability distributions of inputs
    


    


    
## Example: Monty hall

- model the system:

  - car_door: door where car is
  
  - player_choice: door player chose
  
  - strategy: switch or stay
  
  - win if car_door = player_door
  
- assign probabilities to car_door and player_choice

  - both (1/3,1/3,1/3)

- run many iterations

### Some R functions

```{r, eval=FALSE}
# generate variable
variable_name <- 1

# generate vector of size n
n <- 3
vector_name <- numeric(n)

# assign values to vector elements
vector_name[2]<-10
vector_name
```



```{r, eval=FALSE}
# sample k elements from set set_name 
sample(set_name,k)

# example: sample 1 element from the vector (1,2,3)
sample(1:3, 1)
```


### Code (see html version for details)

```{r}
monty_hall_simulation <- function(num_trials = 10000) {
  
  # define variables to save results
  wins_stay <- numeric(num_trials)
  wins_switch <- numeric(num_trials)

  # for loop: repeat num_trials times
  for (i in 1:num_trials) {
    # 1. Host hides the car behind a random door (1, 2, or 3)
    car_door <- sample(1:3, 1)

    # 2. Player chooses a door (1, 2, or 3)
    player_choice <- sample(1:3, 1)

    # 3. staying wins if player_choice is car_door
    wins_stay[i]   <- (player_choice == car_door)
    
    # 4. switching wins if player_choice is not car_door
    wins_switch[i] <- (player_choice != car_door)
  }


  return(list(stay=wins_stay,switch= wins_switch))
}

# Run the simulation (e.g., 1,000 trials)
results <- monty_hall_simulation(num_trials = 10000)
```

### results

```{r, results = "hold"}
cat("Number of trials:", length(results$stay), "\n")
cat("winning by staying:", sum(results$stay), "\n")
cat("winning by switching:", sum(results$switch), "\n")

mean(results$switch)
```


### Python solution

```{python}

import random

def monty_hall_simulation(num_trials=10000):
    """
    Simulates the Monty Hall problem using Monte Carlo method.

    Args:
        num_trials: The number of times to run the simulation.

    Returns:
        A tuple containing the win percentage for staying and switching.
    """
    wins_stay = 0
    wins_switch = 0

    for _ in range(num_trials):
        # Doors are represented by numbers 0, 1, 2

        # 1. Host hides the car behind a random door
        car_door = random.choice([0, 1, 2])

        # 2. Player chooses a door
        player_choice = random.choice([0, 1, 2])

        # 3. staying wins if player_choice is car_door
        if player_choice == car_door:
            wins_stay += 1

        # 4. switching wins if player_choice is not car_door
        if player_choice != car_door:
            wins_switch += 1

    # Calculate win percentages
    win_percentage_stay = (wins_stay / num_trials) * 100
    win_percentage_switch = (wins_switch / num_trials) * 100

    return win_percentage_stay, win_percentage_switch

# Run the simulation
num_simulations = 10000
stay_percentage, switch_percentage = monty_hall_simulation(num_simulations)

print(f"Based on {num_simulations} simulations:")
print(f"  Win percentage by staying: {stay_percentage:.2f}%")
print(f"  Win percentage by switching: {switch_percentage:.2f}%")
```



## Technical summary

- define model $m$ that generates samples of $y$

    - in model $m$ we normally have input variables or parameters
    
    - let's call those inputs $\theta$
    
- define distribution for inputs, $\theta \sim P_\theta$

- one simulation run

  - sample $\theta_1 \sim P_\theta$
  
  - sample $y_1 \sim m(\theta_1)$
  
- $n$ simulation runs result in predictive sample $(y_1, \dots, y_n)$

  - can be understood as sample from probability prediction for $y$
  
## Example Linear regression

- model $m$:
$$y = \beta x + \epsilon$$

- $\theta = (\beta,x,\epsilon)$ where 

    - input variables: $x$ and $\epsilon$
    
    - parameters: $\beta$

- choose input distribution:

  - $x \sim U[0,1]$ (uniform between 0 and 1)
  
  - could come from data distribution
  
  - $\beta,\epsilon \sim \mathcal N(0,1)$
  
  - could come from OLS result
  
- compute predictions $y=m(\beta,x,\epsilon)$

- note: difference to OLS point prediction 

    - $\hat y = \hat \beta x$

### Pseudo code

```{r, eval=F}

for i from 1 to n_replication
  
  # sample inputs
  sample x from uniform
  sample epsilon from standard normal
  sample beta from standard normal
  
  # compute linear model
  compute y[i] = x*beta + epsilon
  
```



### Some R functions

```{r, eval=FALSE}
# sample from normal distribution
rnorm(n = 1, mean = 0, sd = 1)
rnorm(n=1, mean = 2)

# sample from Uniform distribution
runif(n = 1)

# define my own function
my_function <- function(parameter){
  my_result <- parameter*2
  
  return(my_result)
}

# call function
my_function(3)

# replicate function
replicate(n=2, my_function(3))

```
  
### Code simulate

```{r}
# define model
m <- function(beta,x,epsilon){
  y <- x*beta + epsilon
  return(y)
}

# define sampling+model computation
sample1 <- function(){
  
  # sample from input distribution
  x <- runif(n = 1)
  epsilon <- rnorm(n = 1)
  beta <- rnorm(n = 1)
  
  # compute outcome
  y <- m(beta,x,epsilon)
  
  return(y)
}
sample1()
```


### Code: Execute simulation

```{r}
# run simulation function
y_sample <- replicate(n = 1000, sample1())

head(y_sample)
```


### Analyzing the predictive sample

- predictive sample represents probabilistic prediction

    - larger sample size, improves approximation
    
- can look at histogram for illustration

- can look at event probabilities by computing ratios
    $$P(y\le c) \approx \frac 1 n \sum 1_{(y_i \le c)}$$
    
- can derive mean point predictions by computing averages
    $$\hat y \approx \frac 1 n \sum y_i$$
    
- can compute Confidence intervals with quantiles $q_\alpha$ of sample $y_1,\dots,y_n$
    $$CI_{90\%} \approx [q_{5\%}, q_{95\%}]$$

### Some R functions

```{r, eval=FALSE}
# plot histogram
hist(rnorm(100,0,1))

```

### Code: Predictive distribution

```{r}
hist(y_sample)
abline(v=0,col="red",lwd = 5)

mean(y_sample)

mean(y_sample < 1)
```

## Example: Predicting wealth with stock market participation

- core idea:

  - you can invest 1.000 CHF for 10 years
  
  - what is your risk if you put it in a standard stock portfolio?
  
- model: $$y[t+1]=y[t] r[t]$$

- outcome: $y[10]$

- uncertain input is $r[t]$

  - assume that follows normal distribution
  
  - mean and sd based on historic values
  
  
- note: no investment advice, numbers are crude guesstimates.

### Input research (data)

```{r}
library(tidyquant)
library(dplyr)

# Download historical data for the S&P 500 (^GSPC is the Yahoo Finance ticker)
sp500_yearly <- tq_get("^GSPC",
                     from = "1980-01-01",
                     to = Sys.Date()) |>
  group_by(year = lubridate::year(date)) |>
  slice_max(date, n = 1) |>
  ungroup() |> 
  select(year,close)

head(sp500_yearly)
```
### Input research: mean and sd

```{r}
# extract returns
returns <- sp500_yearly |> 
  mutate(r = close/lag(close)-1) |>
  filter(!is.na(r)) |> # keep only complete observations
  pull(r) # pull vector from data.frame

# compute average
mean(returns) 

# compute standard deviation
sd(returns)
  
```

- We will use the assumption $$r[t] \sim_{iid} \mathcal N (0.10, 0.16^2)$$


### Some R functions

```{r, eval=FALSE}
# for loop
for(t in 1:3){ #loop over c(1,2,3)
  # in each loop, t has one of the values of c(1,2,3)
  print(t)
}


```

### Code for one simulation run

```{r, eval=FALSE}
### inputs from analysis above
# average return
average_growth <- 0.10
# sd of return
average_sd <- 0.16

years <- 10

# create vectors
r <- numeric(years)
y <- numeric(years+1)

#starting wealth
y[1] <- 1

# loop over years
for(t in 1:years){
  #sample random return
  r[t] <- rnorm(n = 1,
                mean = 1 + average_growth,
                sd = average_sd)
  # save next time step
  y[t+1]<-y[t]*r[t]
}
y[years+1] #result of model run
```

### Function for one simulation run

```{r}
### inputs from analysis above
sample1 <- function(average_growth = 0.10,
                    average_sd = 0.16,
                    years = 10){
  # create vectors
  r <- numeric(years)
  y <- numeric(years+1)
  
  #starting wealth
  y[1] <- 1
  
  # loop over years
  for(t in 1:years){
    #sample random return
    r[t] <- rnorm(n = 1,
                  mean = 1 + average_growth,
                  sd = average_sd)
    # save next time step
    y[t+1]<-y[t]*r[t]
  }
  return(y[years+1])
}
```

### Run simulation
```{r}
# ensures replicability of random draws
set.seed(1)

# parameters
sim_runs <- 1000

# execute simulation
y <- replicate(sim_runs, sample1())

```


### Results distribution

```{r}
# distribution approximation
hist(y,breaks = 100)
abline(v= mean(y),col="red", lwd=4)
abline(v= 1,col="blue", lwd=4)
```

### Results

- let's compute 

    - expected return (point prediction)
    
    - probability of making losses
    \vspace{1cm}

```{r}
# expected return after 10 years
mean(y)

# probability of loss
mean(y<1)
```


### Python solution

```{python}
# if you use Rstudio, make sure
import numpy as np
import matplotlib.pyplot as plt

# Ensures replicability of random draws
np.random.seed(1)

# Parameters
initial_wealth = 1.0  # Corresponds to w1 in R code
years = 10
sim_runs = 1000

# Inputs from analysis (average annual growth and standard deviation)
average_growth = 0.10
average_sd = 0.16

# Initialize arrays to store results
wealth_paths = np.full((years + 1, sim_runs), np.nan)
# Store initial wealth in the first row
wealth_paths[0, :] = initial_wealth

# Returns array 'r' stores the random growth factor for each year/simulation
# The mean for the normal distribution is 1 + average_growth to represent the factor (e.g., 1.10 for 10% growth)
returns = np.full((years, sim_runs), np.nan)

# Loop over years
for t in range(years):
    # Sample random returns for all simulations for year t
    # Use np.random.normal for sampling from normal distribution
    # The mean is 1 + average_growth because we multiply wealth by this factor
    random_returns = np.random.normal(loc=1 + average_growth, scale=average_sd, size=sim_runs)
    returns[t, :] = random_returns

    # Calculate next year's wealth for all simulations
    wealth_paths[t + 1, :] = wealth_paths[t, :] * returns[t, :]

# Extract final wealth from the last row of the wealth_paths array
final_wealth = wealth_paths[years, :] # Corresponds to y[years+1,] in R

# 1. Distribution Approximation - Plot Histogram
plt.figure(figsize=(10, 6))
plt.hist(final_wealth, bins=100, edgecolor='black') # 'breaks=100' corresponds to bins=100
plt.title(f'Distribution of Wealth after {years} Years ({sim_runs} simulations)')
plt.xlabel('Final Wealth')
plt.ylabel('Frequency')
plt.grid(axis='y', alpha=0.75)
plt.show()

# 2. Probability of Loss
# Calculate the proportion of simulations where final wealth is less than initial wealth
prob_loss = np.mean(final_wealth < initial_wealth) # Corresponds to mean(y[years+1,]<1)

print(f"\nBased on {sim_runs} simulations over {years} years:")
print(f"  Average final wealth: {np.mean(final_wealth):.2f}")
print(f"  Probability of ending with less than initial wealth: {prob_loss:.3f} ({prob_loss*100:.1f}%)")

```


### Critique

- returns vs real returns

- returns follow Gaussian distribution?

```{r,out.width="70%"}
hist(returns)
```


## Summary: Monte-Carlo

Idea:

- define a model 

    - generative (can be simulated from) and 
    
    - probabilistic (is random)
    
- inputs of model have some distribution

- simulate repeatedly from model to generate a sample of predictions

- each simulated outcome represents a probability particle

    - compute probability of event by averaging
    
    - compute mean predictions for real by averaging
    
    - illustrate probabilistic prediction for reals by histogram

# References {.allowframebreaks}
